---
title: "Charbon (Onion Smut) — Data Cleaning"
format:
  html:
    toc: true
    embed-resources: true
---

## Notes

Two data sets are cleaned in this file:

- `donnees_parcelles_2023-25.xlsx` — data collected in the experimental
  parcelles; this is the **training** set used for resampling/tuning.
- `donnees_composites_2023-25.xlsx` — data collected in other (independent)
  fields; this is the **held-out test** set, used only once at the very end
  of the modeling process (see `02_modeling_template.qmd`).

Both must end up with the same predictor columns, names, and types, and the
same outcome-threshold definition, so keep the two cleaning tracks below in
sync. Any step applied to one (renaming, type conversion, threshold cutoffs)
must be mirrored in the other.

## Setup

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(readxl)
library(janitor)
```

## Import — training set (parcelles)

```{r}
#| label: sheets
train_path <- "donnees_parcelles_2023-25.xlsx"
excel_sheets(train_path)
```

```{r}
#| label: import
# TODO: confirm sheet name once excel_sheets() output is known
raw <- read_excel(train_path, sheet = 1)
```

```{r}
#| label: clean-names
df <- raw |> clean_names()
glimpse(df)
```

## Import — test set (composites)

```{r}
#| label: sheets-test
test_path <- "donnees_composites_2023-25.xlsx"
excel_sheets(test_path)
```

```{r}
#| label: import-test
# TODO: confirm sheet name once excel_sheets() output is known
raw_test <- read_excel(test_path, sheet = 1)
```

```{r}
#| label: clean-names-test
df_test <- raw_test |> clean_names()
glimpse(df_test)
```

Compare column names between the two sets before proceeding, since they were
collected independently and may not line up exactly:

```{r}
#| label: compare-columns
setdiff(names(df), names(df_test))
setdiff(names(df_test), names(df))
```

## Initial inspection

Run these on `df` (training) and repeat on `df_test` (composites) as needed.

```{r}
#| label: distinct-counts
df |> summarise(across(everything(), n_distinct))
```

```{r}
#| label: missing-counts
df |> summarise(across(everything(), ~ sum(is.na(.))))
```

```{r}
#| label: distinct-counts-test
df_test |> summarise(across(everything(), n_distinct))
```

```{r}
#| label: missing-counts-test
df_test |> summarise(across(everything(), ~ sum(is.na(.))))
```

```{r}
#| label: categorical-checks
df |> count(annee)
df |> count(cultivar)
df |> count(traitement_de_semences)
df |> count(type)
df |> count(station)
```

```{r}
#| label: categorical-checks-test
df_test |> count(annee)
df_test |> count(variete)
df_test |> count(traitement_de_semence)
df_test |> count(stade)
```

```{r}
#| label: numeric-checks
df |>
  mutate(bin = cut(percent_incidence_moyenne_maximale,
                    breaks = seq(0, 40, by = 2), include.lowest = TRUE)) |>
  count(bin) |>
  arrange(bin)
```

```{r}
#| label: numeric-checks-test
df_test |>
  mutate(bin = cut(incidence_moy,
                    breaks = seq(0, 40, by = 2), include.lowest = TRUE)) |>
  count(bin) |>
  arrange(bin)
```

## Visualize the raw data

```{r}
#| label: viz-incidence-hist
df |>
  ggplot(aes(percent_incidence_moyenne_maximale)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Distribution of maximum incidence (training / parcelles)",
       x = "Percent incidence (moyenne maximale)", y = "Count")
```

```{r}
#| label: viz-incidence-hist-test
df_test |>
  ggplot(aes(incidence_moy)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Distribution of mean incidence (test / composites)",
       x = "Percent incidence (moy)", y = "Count")
```

```{r}
#| label: viz-incidence-by-year
df |>
  ggplot(aes(y = percent_incidence_moyenne_maximale, x = factor(annee))) +
  geom_boxplot() +
  labs(title = "Incidence by year (training)", x = "Year", y = "Percent incidence")
```

```{r}
#| label: viz-incidence-by-cultivar
df |>
  ggplot(aes(y = percent_incidence_moyenne_maximale, x = cultivar)) +
  geom_boxplot() +
  labs(title = "Incidence by cultivar (training)", x = NULL, y = "Percent incidence") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: viz-incidence-by-treatment
df |>
  ggplot(aes(y = percent_incidence_moyenne_maximale, x = traitement_de_semences)) +
  geom_boxplot() +
  labs(title = "Incidence by seed treatment (training)", x = NULL, y = "Percent incidence") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: viz-incidence-vs-spores
df |>
  ggplot(aes(x = spores_g_sol, y = percent_incidence_moyenne_maximale)) +
  geom_point(alpha = 0.5) +
  labs(title = "Incidence vs. soil spore load (training)",
       x = "Spores / g soil", y = "Percent incidence")
```

```{r}
#| label: viz-incidence-vs-temp
df |>
  ggplot(aes(x = semi_se_temp_air, y = percent_incidence_moyenne_maximale)) +
  geom_point(alpha = 0.5) +
  labs(title = "Incidence vs. air temperature (sowing to flag-leaf stage)",
       x = "Mean air temp, sowing→stade étendard", y = "Percent incidence")
```

```{r}
#| label: viz-incidence-vs-rain
df |>
  ggplot(aes(x = semi_se_pluie, y = percent_incidence_moyenne_maximale)) +
  geom_point(alpha = 0.5) +
  labs(title = "Incidence vs. rainfall (sowing to flag-leaf stage)",
       x = "Total rainfall, sowing→stade étendard", y = "Percent incidence")
```

```{r}
#| label: viz-incidence-by-producteur
df |>
  ggplot(aes(y = percent_incidence_moyenne_maximale, x = producteur)) +
  geom_boxplot() +
  labs(title = "Incidence by producer (training)", x = NULL, y = "Percent incidence") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Missing data investigation

```{r}
#| label: missingness-patterns
# Which columns have any missingness, and how much?
df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))
```

```{r}
#| label: missingness-by-year
# Check whether missingness in key columns is concentrated in specific years
# or producers (common when a variable was added/dropped partway through the
# study).
df |>
  group_by(annee) |>
  summarise(across(where(~ any(is.na(.))), ~ sum(is.na(.))))
```

```{r}
#| label: missingness-sample-rows
df |> filter(if_any(everything(), is.na)) |> slice_sample(n = min(10, nrow(df)))
```

## Type conversions / renaming

Apply identical conversions/renames to both `df` and `df_test`.

```{r}
#| label: type-conversions
# TODO: convert dates, coerce categorical columns to factors, rename columns
# for clarity as needed.
```

```{r}
#| label: type-conversions-test
# TODO: mirror the conversions above for df_test.
```

## Define the outcome variables

Rather than one multi-class outcome, we build **five separate binary
outcomes**, one per threshold (1%, 2%, 3%, 4%, 5%). Each is analyzed as its
own classification problem in its own copy of `02_modeling_template.qmd`
(e.g. `02_modeling_threshold_1.qmd`, ... `02_modeling_threshold_5.qmd`), so a
row can be, e.g., a positive case for `outcome_1` (incidence ≥ 1%) while also
being negative for `outcome_5` (incidence < 5%) — these are not mutually
exclusive levels of one factor.

The training data has a continuous maximum-incidence column
(`percent_incidence_moyenne_maximale`) and also happens to contain
pre-computed indicator columns (`seuil_1_percent_incidence` ...
`seuil_5_percent`). We calculate the five binary outcomes directly from the
continuous column ourselves, rather than relying on the pre-computed
indicators, so the derivation is transparent and uses the identical logic as
the composites set below. (Worth spot-checking the derived columns against
the existing `seuil_*` columns as a sanity check — see below.)

```{r}
#| label: outcome-thresholds-train
# TODO: confirm >= is the intended boundary (vs strictly >).
df <- df |>
  mutate(
    outcome_1 = factor(if_else(percent_incidence_moyenne_maximale >= 1, "above", "below"), levels = c("below", "above")),
    outcome_2 = factor(if_else(percent_incidence_moyenne_maximale >= 2, "above", "below"), levels = c("below", "above")),
    outcome_3 = factor(if_else(percent_incidence_moyenne_maximale >= 3, "above", "below"), levels = c("below", "above")),
    outcome_4 = factor(if_else(percent_incidence_moyenne_maximale >= 4, "above", "below"), levels = c("below", "above")),
    outcome_5 = factor(if_else(percent_incidence_moyenne_maximale >= 5, "above", "below"), levels = c("below", "above"))
  )
```

```{r}
#| label: outcome-thresholds-sanity-check
# Sanity check: compare our derived outcomes against the pre-existing
# seuil_* indicator columns — mismatches would suggest a different boundary
# convention (>= vs >) or a discrepancy in how "maximale" incidence was
# computed upstream.
df |>
  summarise(
    mismatch_1 = sum((outcome_1 == "above") != (seuil_1_percent_incidence == 1)),
    mismatch_2 = sum((outcome_2 == "above") != (seuil_2_percent_incidence == 1)),
    mismatch_3 = sum((outcome_3 == "above") != (seuil_3_percent == 1)),
    mismatch_4 = sum((outcome_4 == "above") != (seuil_4_percent == 1)),
    mismatch_5 = sum((outcome_5 == "above") != (seuil_5_percent == 1))
  )
```

```{r}
#| label: outcome-thresholds-test
# Composites set only has a continuous incidence column; derive the same
# five binary indicators directly from it.
# TODO: confirm `incidence_moy` is on the same 0-100 percent scale as the
# training incidence column, and whether >= or > is the intended boundary.
df_test <- df_test |>
  mutate(
    outcome_1 = factor(if_else(incidence_moy >= 1, "above", "below"), levels = c("below", "above")),
    outcome_2 = factor(if_else(incidence_moy >= 2, "above", "below"), levels = c("below", "above")),
    outcome_3 = factor(if_else(incidence_moy >= 3, "above", "below"), levels = c("below", "above")),
    outcome_4 = factor(if_else(incidence_moy >= 4, "above", "below"), levels = c("below", "above")),
    outcome_5 = factor(if_else(incidence_moy >= 5, "above", "below"), levels = c("below", "above"))
  )
```

```{r}
#| label: outcome-balance
df |> summarise(across(starts_with("outcome_"), ~ mean(.x == "above")))
df_test |> summarise(across(starts_with("outcome_"), ~ mean(.x == "above")))
```

Class balance in the training data (595 rows): threshold 1% ≈ 38% positive,
2% ≈ 28%, 3% ≈ 15%, 4% ≈ 6%, 5% ≈ 4%. Imbalance increases sharply at the
higher thresholds — the 4% and 5% analyses have few positive cases (38 and
21 out of 595), which will make 10-fold CV noisy for those two models
specifically (some folds may end up with very few or zero positive cases).
Consider fewer folds (e.g. `v = 5`) or repeated CV for those two thresholds,
and check the composites set's class balance too, since it is much smaller
(64 rows) and may not contain any positive cases at the higher thresholds.

## Save cleaned data

Both `df` and `df_test` now carry all five `outcome_1` ... `outcome_5`
columns, both derived directly from each set's continuous incidence column
(`percent_incidence_moyenne_maximale` / `incidence_moy`) rather than from
pre-computed indicators. Each modeling template picks one outcome at a time
via a single `outcome_col` setting, and should drop both the raw incidence
column and the (now redundant) `seuil_*` indicator columns as potential
leakage.

```{r}
#| label: save
# TODO: confirm final column set before saving; keep column names/types
# identical between the two files.
write_rds(df, "donnees_parcelles_clean.rds")
write_rds(df_test, "donnees_composites_clean.rds")
```
