---
title: "Charbon (Onion Smut) — Modeling Template"
format:
  html:
    toc: true
    embed-resources: true
---

# Notes

This template covers ONE model type × ONE threshold at a time (e.g. decision
tree for the 1% threshold). Since we're running five separate binary
analyses (1%, 2%, 3%, 4%, 5% thresholds) × three model types, duplicate this
file per combination (e.g. `02_tree_threshold1.qmd`, `02_rf_threshold1.qmd`,
..., `02_boost_threshold5.qmd`), keeping the recipe/resampling structure
identical across copies so results are comparable across thresholds and
models. Alternatively, once the pipeline is working for one threshold,
consider combining the three model specs with `workflowsets::workflow_set()`
+ `workflow_map()` for a given threshold, or wrapping the whole file in a
function/loop over `outcome_col` if you'd rather iterate programmatically
than duplicate files.

**Select which threshold this copy analyzes** by setting `outcome_col`
below — this is the only thing that needs to change between the five
threshold copies of this file (assuming predictors are otherwise identical).

## Setup

```{r}
#| label: setup
#| message: false
# `tidymodels` is a meta-package: loading it attaches the packages we need
# for the whole pipeline — `recipes` (preprocessing), `parsnip` (unified
# model specification syntax), `rsample` (splitting/resampling), `tune` +
# `dials` (hyperparameter tuning), `workflows` (bundling recipe + model), and
# `yardstick` (performance metrics).
library(tidymodels)
library(themis)   # subsampling steps for class imbalance (step_downsample, step_upsample, step_smote)
library(vip)       # variable importance plots for the final fitted model
# library(bonsai)  # only needed if using lightgbm engine for boosted trees

# Several tidyverse packages (e.g. dplyr, purrr) export functions with the
# same name as tidymodels functions (e.g. `slice()`). tidymodels_prefer()
# resolves these conflicts by making the tidymodels version the default
# whenever both are loaded.
tidymodels_prefer()
```

## Load cleaned data

Training data (`parcelles`) and the independent test data (`composites`) were
cleaned in `01_data_cleaning.qmd` and saved separately, each carrying five
binary outcome columns (`outcome_1` ... `outcome_5`, one per threshold). The
composites data is a genuinely independent field test set — collected in
different fields than the training/tuning data — so it is **not** produced
via `initial_split()`; it is loaded as-is and touched only once, at the very
end.

```{r}
#| label: load-data
# read_rds() restores the exact R objects (types, factor levels, etc.) saved
# by 01_data_cleaning.qmd — safer than re-reading the Excel files here, since
# we want the cleaning decisions made in that file to be the single source of
# truth for both the training and test data.
train_full <- read_rds("donnees_parcelles_clean.rds")
test_full  <- read_rds("donnees_composites_clean.rds")

# Each cleaned data set carries five candidate outcome columns
# (outcome_1 ... outcome_5), one binary factor per incidence threshold.
# This template analyzes ONE threshold at a time, so we pick a single
# column here and rename it to the generic name `outcome` — that way the
# rest of the pipeline below (recipe, model formula, metrics) doesn't need
# to change between the five copies of this file; only this line does.
# TODO: set which threshold this copy of the template analyzes.
outcome_col <- "outcome_1"

# Besides the four *other* outcome_* columns (which must be dropped so they
# aren't accidentally used as predictors for this threshold), we also drop
# any column that was used to derive the outcomes in the first place. If we
# kept, say, percent_incidence_moyenne_maximale as a predictor, the model
# could learn the trivial rule "incidence >= 1 -> above" instead of learning
# from the actual agronomic/weather predictors — this is called "leakage"
# and would make the model's test performance look artificially perfect.
# TODO: replace with the actual predictor set once finalized.
leakage_cols <- c("seuil_1_percent_incidence", "seuil_2_percent_incidence",
                   "seuil_3_percent", "seuil_4_percent", "seuil_5_percent",
                   "percent_incidence_moyenne_maximale", "incidence_moy")

# rename(outcome = all_of(outcome_col)) renames whichever column outcome_col
# points to (e.g. "outcome_1") to the literal name `outcome`.
# select(-starts_with("outcome_"), -any_of(leakage_cols), outcome) drops the
# remaining outcome_* columns and the leakage columns, then re-adds `outcome`
# at the end (select() drops it too, since it also starts with "outcome_",
# so we have to explicitly list it again to keep it).
train <- train_full |>
  rename(outcome = all_of(outcome_col)) |>
  select(-starts_with("outcome_"), -any_of(leakage_cols), outcome)

test <- test_full |>
  rename(outcome = all_of(outcome_col)) |>
  select(-starts_with("outcome_"), -any_of(leakage_cols), outcome)
```

## Resampling for tuning

All hyperparameter tuning and model comparison happens via cross-validation
on `train` only. `test` (composites) is never used until `last_fit()` below.

For the higher thresholds (4%, 5%) the positive class is rare (~6% and ~4%
of training rows respectively) — consider fewer folds (e.g. `v = 5`) or
`vfold_cv(..., repeats = ...)` for those two threshold copies, since 10 folds
may leave some folds with very few positive cases.

```{r}
#| label: resamples
# set.seed() makes the random fold assignment reproducible — anyone re-running
# this chunk gets the exact same folds. Any seed works; what matters is
# fixing *some* seed before any random process (splitting, subsampling,
# hyperparameter grid sampling, etc.).
set.seed(6182)

# vfold_cv() implements v-fold (a.k.a. k-fold) cross-validation: it splits
# `train` into v roughly equal folds. For each of the v tuning iterations,
# v-1 folds are used to fit the model ("analysis" set) and the remaining
# fold is used to evaluate it ("assessment" set) — this repeats v times so
# every row gets used for assessment exactly once. Averaging performance
# across the v folds gives a more stable estimate of how well a given set of
# hyperparameters generalizes than a single train/validation split would.
# `strata = outcome` ensures each fold has approximately the same proportion
# of "above"/"below" cases as the full training set — important here because
# the outcome is imbalanced, especially at higher thresholds.
folds <- vfold_cv(train, v = 10, strata = outcome)  # TODO: reduce v for rarer thresholds
```

## Recipe

```{r}
#| label: recipe
# A `recipe` is tidymodels' preprocessing specification: a sequence of steps
# applied to the data before it reaches the model. Critically, when a recipe
# is used inside a workflow with resampling (as we do below), each step is
# re-estimated on only the analysis (training) fold and then applied to the
# assessment fold — this avoids leaking information from validation data
# into preprocessing decisions (e.g. computing a mean to normalize by should
# only ever use training rows).
#
# recipe(outcome ~ ., data = train) declares the model formula: predict
# `outcome` from every other column in `train` ("."). The `data` argument is
# only used to record column names/types, not to fit anything yet.
rec <- recipe(outcome ~ ., data = train) |>
  # step_novel() reserves a "new" factor level for any category that shows
  # up in future data (e.g. the composites test set, or new fields) but
  # never appeared in the training data — without this, applying the recipe
  # to unseen categories would error out.
  step_novel(all_nominal_predictors()) |>
  # step_dummy() converts categorical (factor/character) predictors into
  # 0/1 dummy columns (one-hot encoding minus a reference level), since most
  # engines here (rpart, ranger, xgboost) either require numeric input
  # (xgboost) or benefit from consistent encoding across engines.
  step_dummy(all_nominal_predictors()) |>
  # step_zv() drops "zero-variance" columns — predictors that take only one
  # value in the training data (e.g. a dummy column that ended up constant
  # after filtering) and so carry no information.
  step_zv(all_predictors()) |>
  # step_normalize() centers and scales numeric predictors to mean 0 / sd 1.
  # Tree-based models (all three we use here) are invariant to monotonic
  # rescaling of predictors, so this step doesn't change tree-based model
  # performance — it's included mostly as a placeholder/reminder in case a
  # non-tree model is added later; feel free to remove it for a pure
  # tree/forest/boosting comparison.
  step_normalize(all_numeric_predictors()) |>  # optional; trees don't require this
  # TODO: adjust predictors, encoding steps, and imbalance handling as needed.
  # Only keep ONE of the step_*sample() lines below — these correct for
  # class imbalance by resampling the *training* data seen during model
  # fitting (never the assessment/test data, which must reflect real-world
  # class proportions for honest evaluation):
  #   step_downsample(outcome) — randomly removes majority-class ("below")
  #     rows until both classes are equally represented. Simple, but throws
  #     away data — costly here given the small sample size (595 rows),
  #     especially at the rarer 4%/5% thresholds.
  #   step_upsample(outcome)  — randomly duplicates minority-class ("above")
  #     rows until both classes are equal. Keeps all original data but can
  #     make the model overfit to the duplicated minority rows.
  #   step_smote(outcome)     — synthesizes new minority-class rows by
  #     interpolating between existing minority-class neighbors (SMOTE),
  #     rather than duplicating them exactly. Needs enough minority
  #     neighbors to work well, which is a concern for the rarest thresholds.
  # For the 4%/5% thresholds (rare positive class), downsampling discards a
  # lot of majority-class data given the small sample size — upsampling or
  # SMOTE may be preferable there; compare against downsampling empirically
  # rather than assuming.
  # step_downsample(outcome)
  # step_upsample(outcome)
  # step_smote(outcome)
  identity()  # placeholder so the pipe chain above is syntactically valid with steps commented out
```

## Model specification

Keep only the block relevant to this copy of the template. Each `parsnip`
model spec below declares which hyperparameters to tune (via `tune()`,
a placeholder telling tidymodels "search over values for this argument
rather than using a fixed value") and which underlying R engine/package
implements the model.

```{r}
#| label: model-spec-tree
# A single decision tree recursively splits the data on predictor values to
# separate the two outcome classes as purely as possible at each split.
# Simple and interpretable, but a single tree tends to overfit training data
# unless its growth is controlled by the hyperparameters below:
#   cost_complexity — penalty added per additional split/leaf ("cp" in
#     rpart); larger values prune the tree more aggressively (simpler tree).
#   tree_depth      — maximum number of splits from root to any leaf.
#   min_n           — minimum number of observations required in a node
#     before it can be split further.
tree_spec <- decision_tree(
  cost_complexity = tune(),
  tree_depth      = tune(),
  min_n           = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")
```

```{r}
#| label: model-spec-rf
# Random forest fits many decision trees (the `trees` argument) on
# bootstrap-resampled versions of the training data, each considering only a
# random subset of predictors (`mtry`) at every split. Averaging (technically,
# majority-voting for classification) across many decorrelated trees reduces
# the overfitting a single tree is prone to.
#   mtry  — number of predictors randomly sampled as candidates at each
#     split (smaller values decorrelate trees more, at the cost of each
#     individual tree being weaker).
#   trees — number of trees in the forest.
#   min_n — minimum observations required in a node before splitting
#     (controls how deep/complex individual trees can grow).
rf_spec <- rand_forest(
  mtry  = tune(),
  trees = tune(),
  min_n = tune()
) |>
  # importance = "impurity" tells ranger to record, for each predictor, how
  # much it reduces node impurity across all trees on average — this is
  # what powers the vip() variable-importance plot at the end of this file.
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")
```

```{r}
#| label: model-spec-boost
# Boosted trees (here, xgboost) build trees sequentially rather than
# independently: each new tree is fit to correct the errors of the trees
# fit so far, weighted by a learning rate. This often achieves the best raw
# predictive performance of the three methods, at the cost of more
# hyperparameters to tune and less interpretability than a single tree.
#   mtry           — predictors randomly sampled as candidates per split.
#   trees          — number of sequential boosting rounds (trees added).
#   min_n          — minimum observations required in a node to split.
#   tree_depth     — maximum depth of each individual tree (kept shallow in
#     boosting, since trees are meant to be "weak learners" corrected
#     iteratively, not strong learners on their own).
#   learn_rate     — how much each new tree's predictions are down-weighted
#     before being added to the ensemble; smaller values need more trees but
#     generalize better (less prone to overfitting).
#   loss_reduction — minimum improvement in the loss function required to
#     make an additional split (a regularization term, "gamma" in xgboost).
#   sample_size    — proportion of training rows randomly subsampled to
#     grow each tree (stochastic gradient boosting; adds regularization
#     similar to bagging in random forests).
boost_spec <- boost_tree(
  mtry          = tune(),
  trees         = tune(),
  min_n         = tune(),
  tree_depth    = tune(),
  learn_rate    = tune(),
  loss_reduction = tune(),
  sample_size   = tune()
) |>
  set_engine("xgboost") |>
  set_mode("classification")
```

```{r}
#| label: select-spec
# TODO: pick the spec matching this template copy (tree_spec, rf_spec, or
# boost_spec). Everything downstream (workflow, tuning, evaluation) refers
# only to `model_spec`, so this is the only line that needs to change
# between the tree/RF/boosted copies of this file.
model_spec <- tree_spec  # or rf_spec / boost_spec
```

## Workflow

```{r}
#| label: workflow
# A `workflow` bundles the preprocessing recipe and the model specification
# into one object, so that both are fit/applied together — this ensures the
# exact same preprocessing (fit on the analysis fold) is used whenever the
# model is fit or predictions are made, and lets `tune_grid()`/`last_fit()`
# below operate on a single object instead of juggling recipe and model
# separately.
wf <- workflow() |>
  add_recipe(rec) |>
  add_model(model_spec)
```

## Tuning grid

```{r}
#| label: tuning-grid
# TODO: mtry's valid range depends on the number of predictors available
# after the recipe runs (it can't sample more predictors than exist), which
# tidymodels doesn't know automatically — finalize() fills in that range by
# inspecting `train`. Needed whenever mtry (rf_spec, boost_spec) is tuned;
# not needed for tree_spec, which has no mtry parameter.
# params <- extract_parameter_set_dials(wf) |>
#   finalize(train)

# grid_latin_hypercube() generates a space-filling random sample of
# hyperparameter combinations (a Latin hypercube design spreads points
# evenly across the multi-dimensional parameter space, unlike simple random
# sampling which can cluster points). This is generally more efficient than
# a full grid search (grid_regular()) when tuning several hyperparameters
# at once, since a regular grid's size grows multiplicatively with each
# added parameter.
# extract_parameter_set_dials(wf) reads off the tunable parameters and their
# default ranges directly from `wf` (the tree_spec/rf_spec/boost_spec chosen
# above), so this line doesn't need to change between model copies of this
# template — it automatically adapts to whichever spec is active.
grid <- grid_latin_hypercube(
  extract_parameter_set_dials(wf),
  size = 25  # TODO: adjust grid size — more candidate combinations gives a
             # finer search but takes proportionally longer to run
)
```

## Tune

```{r}
#| label: tune
#| message: false
# metric_set() bundles several yardstick metrics to compute together during
# tuning, so we don't need separate tune_grid() calls per metric:
#   roc_auc     — area under the ROC curve; overall ability to rank "above"
#     cases higher than "below" cases across all possible classification
#     thresholds (probability cutoffs), independent of class imbalance.
#   pr_auc      — area under the precision-recall curve; often more
#     informative than roc_auc when the positive class is rare (as at the
#     4%/5% thresholds here), since it focuses on how well positives are
#     identified without being diluted by the (large) number of true
#     negatives.
#   sensitivity — proportion of true "above" cases correctly identified
#     (a.k.a. recall / true positive rate).
#   specificity — proportion of true "below" cases correctly identified
#     (true negative rate).
#   f_meas      — harmonic mean of precision and sensitivity; a single
#     summary number balancing both types of error.
metrics <- metric_set(roc_auc, pr_auc, sensitivity, specificity, f_meas)

set.seed(8524)
# tune_grid() is the workhorse of the tuning process: for every row of
# `grid` (i.e. every candidate hyperparameter combination), it fits the
# workflow on each fold's analysis set and evaluates it on that fold's
# assessment set, recording all requested `metrics`. With 10 folds and a
# grid of 25 combinations, this fits 250 models in total.
tune_res <- tune_grid(
  wf,
  resamples = folds,
  grid = grid,
  metrics = metrics,
  # save_pred = TRUE keeps the per-row predictions from every fold/parameter
  # combination, which lets us inspect them later if needed (at the cost of
  # more memory) — not strictly required, but convenient during exploration.
  control = control_grid(save_pred = TRUE)
)
```

```{r}
#| label: tuning-results
# autoplot() visualizes how each metric varies across the values tried for
# each hyperparameter — useful for spotting whether performance is fairly
# flat (many hyperparameter values work about as well) or sharply peaked
# (only a narrow range of values performs well) across the search space.
autoplot(tune_res)

# show_best() lists the top hyperparameter combinations (by default, 5)
# ranked by the chosen metric, averaged across the 10 folds — this is what
# we inspect to decide which single combination to carry forward.
show_best(tune_res, metric = "roc_auc")
```

## Finalize and fit

```{r}
#| label: finalize
# select_best() picks the single best-performing hyperparameter combination
# (by the chosen metric, averaged across folds) from the tuning results.
best_params <- select_best(tune_res, metric = "roc_auc")  # TODO: confirm metric

# finalize_workflow() substitutes those specific hyperparameter values into
# the workflow's model spec, replacing the tune() placeholders — the result
# is an ordinary (non-tunable) workflow ready to be fit on data.
final_wf <- finalize_workflow(wf, best_params)
```

```{r}
#| label: manual-split
# last_fit() (below) expects an `rsplit` object — the same kind of object
# initial_split() would normally produce — bundling a training portion
# ("analysis") and a test portion ("assessment") together. We build one
# manually from the two separately-loaded data sets rather than from
# initial_split(), since train/test here come from different fields (not a
# random split of one data set) and we want to preserve that real-world
# separation rather than re-shuffling rows across the two sources.
split <- make_splits(list(analysis = train, assessment = test))
```

```{r}
#| label: last-fit
# last_fit() does two things in one call: (1) fits `final_wf` (recipe +
# finalized model) on the analysis set (all of `train`), and (2) evaluates
# the fitted model once on the assessment set (all of `test`, the composites
# data). This is the ONLY time the composites data touches the model
# fitting/selection process — everything above (tuning, hyperparameter
# selection) used only `train` via cross-validation, so this final number is
# an honest estimate of performance on genuinely new fields.
final_fit <- last_fit(final_wf, split, metrics = metrics)
collect_metrics(final_fit)
```

## Evaluate on test set

```{r}
#| label: confusion-matrix
# collect_predictions() extracts the row-level predicted class (.pred_class)
# and predicted probabilities for each class from final_fit, for every row
# of the composites (test) set.
# conf_mat() cross-tabulates true outcome vs. predicted class — the
# diagonal counts correct predictions; off-diagonal cells show which kind of
# error (false positive vs. false negative) the model makes, and how often.
collect_predictions(final_fit) |>
  conf_mat(truth = outcome, estimate = .pred_class)
```

```{r}
#| label: roc-curve
# roc_curve() traces sensitivity vs. (1 - specificity) as the probability
# cutoff for classifying a row as "above" is swept from 0 to 1 — this shows
# the full tradeoff between catching true "above" cases and incorrectly
# flagging "below" cases as the decision threshold changes, rather than
# collapsing that tradeoff into the single default 0.5 cutoff used by
# .pred_class above.
collect_predictions(final_fit) |>
  roc_curve(outcome, .pred_above) |>  # TODO: confirm positive-class column name
  autoplot()
```

## Variable importance

```{r}
#| label: vip
# extract_workflow() pulls the fitted workflow object out of final_fit, and
# extract_fit_parsnip() pulls the underlying fitted model object out of that
# workflow (e.g. the rpart/ranger/xgboost model object itself).
# vip() then plots each predictor's importance to the model's predictions —
# for rpart/xgboost this is a built-in importance measure from the engine;
# for ranger it uses the impurity-based importance we requested via
# importance = "impurity" in the model spec above. This helps identify which
# agronomic/weather variables the model is actually relying on, which is
# often as valuable as the predictive performance itself for this kind of
# study.
extract_workflow(final_fit) |>
  extract_fit_parsnip() |>
  vip()
```
